
---
title: "Tutorial: Bayesian Pliable Lasso with Horseshoe Prior"
author: "The Tien Mai"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial: hspliable}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The **`hspliable`** package provides **Bayesian inference for pliable lasso models** using the **Horseshoe prior**.
The pliable lasso framework extends standard sparse regression by allowing **interaction effects between predictors and modifying variables**:

* Sparse *main effects*
* Sparse *interaction effects*
* A *hierarchical structure* that encourages meaningful interactions
* Extension to **Gaussian**, **Logistic**, and **Poisson** GLMs
* Support for **missing responses**

This vignette demonstrates how to:

1. Install the package
2. Fit pliable Horseshoe models
3. Inspect posterior samples and parameter estimates
4. Use the linear, logistic, and Poisson model variants

The underlying methodology is described in:

> **“Bayesian Pliable Lasso with Horseshoe Prior for Interaction Effects in GLMs with Missing Responses.”**

---

# Installation

```r
devtools::install_github("tienmt/hspliable")
library(hspliable)
```

---

# 1. Linear Pliable Model Example

We first illustrate the linear pliable lasso with simulated data.

```r
library(hspliable)

# simulate data
ntest <- 50
n  <- 200 
p  <- 10 
q  <- 2

xx <- matrix(rnorm((n + ntest) * p), (n + ntest), p)
X <- xx[1:n, ]
xtest <- xx[-(1:n), ]

zz <- matrix(rnorm((n + ntest) * q), (n + ntest), q)
Z <- zz[1:n, ]
ztest <- zz[-(1:n), ]

beta_true <- c(2, -2, 2, 2, rep(0, p - 4))

theta_true <- matrix(0, p, q)
theta_true[1:3, ] <- matrix(
  c(rep(1, q),
    rep(-2, q),
    c(1:q)),
  3, q, byrow = TRUE
)

theta0_true <- 0.5
beta0_true  <- -1

yy <- beta0_true +
      zz %*% rep(theta0_true, q) +
      rowSums(sapply(1:p, function(j)
        xx[, j] * (beta_true[j] + zz %*% theta_true[j, ])
      )) +
      rnorm(n + ntest)

y      <- yy[1:n]
ytest  <- yy[-(1:n)]
```

### Fit the pliable Horseshoe regression model

```r
fit_pHS <- pliable_HS(y, X, Z, n_iter = 1000, burn_in = 500)
```

Posterior means for the main effects:

```r
(be_HS_2 <- colMeans(fit_pHS$beta))
beta_true
```

Posterior means for interaction effects:

```r
(theta_HS_2 <- apply(fit_pHS$theta, c(1, 2), mean))
theta_true
```

---

# 2. Logistic Pliable Model

We now illustrate a sparse logistic pliable regression model.

```r
library(hspliable)

ntest <- 500
n  <- 200 
p  <- 10
q  <- 1

xx <- matrix(rnorm((n + ntest) * p), (n + ntest), p)
X  <- xx[1:n, ]
xtest <- xx[-(1:n), ]

zz <- matrix(rnorm((n + ntest) * q), (n + ntest), q)
Z  <- zz[1:n, ]
ztest <- zz[-(1:n), ]

beta_true <- c(2, -2, 2, 2, rep(0, p - 4))
theta_true <- matrix(0, p, q)

theta0_true <- 0.5
beta0_true  <- 2

# linear predictor
my_mu <- beta0_true +
         zz %*% rep(theta0_true, q) +
         rowSums(sapply(1:p, function(j)
           xx[, j] * (beta_true[j] + zz %*% theta_true[j, ])
         ))

yy <- rbinom(n + ntest, 1, prob = boot::inv.logit(my_mu))
y <- yy[1:n]
```

### Fit the logistic pliable model

```r
fit2_cpp <- pliable_HS_logistic(
  y, X, as.matrix(Z),
  eps       = 1e-10,
  n_iter    = 10000,
  burn_in   = 3000,
  clamp_min = 1e-15,
  verbose   = TRUE
)
```

Parameter estimates:

```r
b_pHS_cpp <- colMeans(fit2_cpp$beta)
theta_pHS <- apply(fit2_cpp$theta, c(2, 3), mean)

b_pHS_cpp
beta_true

theta_pHS
theta_true
```

Mean squared error summaries:

```r
cat("MSE(beta) =", sum((b_pHS_cpp - beta_true)^2))
cat("MSE(theta) =", sum((theta_pHS - theta_true)^2))
```

---

# 3. Poisson Pliable Model

```r
library(hspliable)
set.seed(1)

n <- 100
p <- 10
q <- 2

X <- matrix(rnorm(n * p), n, p)
Z <- matrix(rnorm(n * q), n, q)

beta_true  <- c(2, -2, 0, 2, rep(0, p - 4)) / 4
theta_true <- matrix(0, p, q)
theta_true[1:2, ] <- matrix(
  c(rep(1, q),
    rep(-2, q),
    c(1:q)),
  2, q, byrow = TRUE
) / 4

theta0_true <- rep(0.5, q)
beta0_true  <- 2

eta <- beta0_true +
       Z %*% theta0_true +
       rowSums(sapply(1:p, function(j)
         X[, j] * (beta_true[j] + Z %*% theta_true[j, ])
       ))

y <- rpois(n, lambda = exp(eta))
```

### Fit the Poisson pliable model

```r
fit <- pliable_HS_poisson(y, X, Z,
                          n_iter = 5000L,
                          burn_in = 1000L,
                          verbose = TRUE)
```

Posterior means:

```r
colMeans(fit$beta)
beta_true

apply(fit$theta, c(2, 3), mean)
```

---

# Summary

The **`hspliable`** package provides:

* Bayesian estimation of sparse pliable regression models
* Horseshoe priors for enhanced shrinkage
* Unified handling of **Gaussian**, **Logistic**, and **Poisson** responses
* Support for **missing Y values**
* Efficient C++ implementations for scalable MCMC sampling

This vignette demonstrated the essential workflow for fitting pliable models and extracting posterior summaries.

For additional details or feature requests, please visit:
**[https://github.com/tienmt/hspliable](https://github.com/tienmt/hspliable)**

---

# Session Info

```{r}
sessionInfo()
```

```
